Correcting for bias in high throughput sequencing data

### 2011-06-14 seqbias-paper.tex
more revisions

To estimate sequencing bias, we train a Bayesian network classifier in which
each node represents a position in the sequence, relative to the read start, and
edges encode dependency between positions.  Bayesian networks have been applied
to recognize motifs in nucleotide sequences in the past, in particular in
modeling splice cites \citep{Cai2000, Chen2005} and transcription factor binding
sites \citep{Ben-Gal2005, Grau2006, Pudimat2005}.

@@

To begin, we generate $n$ training examples, where $n = 200,000$ is typical, half of
which labeled as foreground, the other background. To do so, we sample the
positions of $n/2$ reads. Duplicate reads are ignored, to avoid overfitting to
any exceedingly highly expressed locus.  For each foreground position $i$
sampled, we then sample a background position by drawing a random offset $\delta
\sim \text{Normal}( 0, \sigma )$, where by default $\sigma = 50$ is chosen, and
taking $i' = \lceil i + \delta \rceil$. By using such a scheme we attempt to
mitigate the effects of biological sequence bias, sampling positions that are
more likely to be biologically similar.

This sampling procedure produces a set of $n$ genomic positions with
accompanying labels $T = \{ (i_1,x_1), (i_2,x_2), \dots, (i_n,x_n) \}$. The
label $x_k$ is binary, indicating classification as background ($x_{i_k} =
0$) or foreground ($x_{i_k} = 1$). We will denote the nucleotide sequence at
position $i_k$ as $s_{i_k}$.

%% TODO: leaving it here for now
%% -------------------------------------------------------

@@

To determine the structure and parameters of the Bayesian network, we use a
hill-climbing approach similar to BNC algorithm \cite{Grossman2004}.
The network structure is determined by greedily optimizing the conditional
log-likelihood.
\begin{align*}
\ell &= \sum_{k=1}^{n} \log \Pr( x_{i_k} | s_{i_k} ) \\
&=
\sum_{k=1}^{n} \log \frac{ \Pr(s_{i_k} | x_{i_k}) \Pr( x_{i_k} ) }{
\sum_{x \in \{0,1\}} \Pr( s_{i_k} | x ) \Pr(x) } \\
%&= \sum_{k=1}^{n} \log \frac{ \Pr( s_{i_k} | x_{i_k} ) }
%{ \sum_{x \in \{0,1\}} \Pr( s_{i_k} | x ) }
\end{align*}

@@

As we will be estimating parameters and evaluating the likelihood on the same
set of samples, simply maximizing the likelihood would severely overfit the
training set. We thus penalize model complexity using the Akaike information
criterion with the second order sample size correction \cite{Hurvich1989}.
Where $m$ is the number of parameters needed to specify the model, we maximize,
$$ \ell' = \ell - m - \frac{(m + 1)(m + 2)}{n - m - 2} $$

At each step of the optimization procedure, every possible edge or position
addition, removal, or edge reversal that produces a valid network (i.e. does not
introduce a cycle) is evaluated, and the alteration that increases the score
$\ell'$ the most is kept.  This process is repeated until a local maxima is
found, in which no single alteration to the network will increase the score.

Given the network structure, the parameters are estimated directly from the
observed nucleotide frequencies. This choice of parameters maximizes the
likelihood of the model, but not necessarily the conditional likelihood. They
are equivalent asymptotically, thus, as per the BNC algorithm, we adopt this
approximation rather than  resorting to numerical methods which might render the
algorithm too slow to be practical.

The run time of the training procedure is further reduced in practice by imposing the
following two restrictions on the structure of the network,
\begin{enumerate}
\item The in-degree (i.e. number of parents) of any node must be less than some
number $p_{\text{max}}$.
\item $|j - i| \le d_{\text{max}}$ for all edges $(i,j)$ and some number $d_{\text{max}}$.
\end{enumerate}
The second of these encodes the assumption that distant nucleotides are
effectively independent.

Figure \ref{fig:structure} shows an example of the structure learned when this
procedure is applied to 200,000 reads from the Trapnell data set
\cite{Trapnell2010}. The structure learned is non-trivial, and
positions that uninformative are excluded automatically.

@@

Because we cannot observe directly the underlying RNA-abundance, our evaluation
strategy relies on testing two assumptions we make of an ideal, unbiased experiment.
\begin{enumerate}
\item Positional nucleotide frequencies (as in Figure \ref{fig:freqs}) measured
from reads within exons should not differ greatly from frequencies measured by
sampling uniformly within the same exons.
\item Read counts across exons should follow approximately a Poisson process.
\end{enumerate}

We applied our procedure (labeled ``BN'') as well as those of Li, et. al, and
Hansen, et. al., which are implemented in the R packages \texttt{mseq} and
\texttt{Genominator}, respectively. The procedures were applied to four publicly
available data sets \cite{Bullard2010} \cite{Mortazavi2008} \cite{Trapnell2010}
\cite{Wetterbom2010}, as well as an unpublished data set of our own (see, Table
\ref{tab:datasets}).

%% TODO
\begin{comment}
\begin{tablehere}
\begin{center}
\footnotesize{
\begin{tabular}{lllr}
\textbf{Experiment} & \textbf{Species} & \textbf{Platform} & \textbf{Read Length} \\ \hline
Wetterbom \cite{Wetterbom2010} & Chimpanzee & ABI & 33 \\
Katze (unpublished) & Macaque & ABI & 50 \\
Bullard \cite{Bullard2010} & Human & Illumina & 35 \\
Mortazavi \cite{Mortazavi2008} & Mouse & Illumina & 33 \\
Trapnell \cite{Trapnell2010} & Mouse & Illumina & 75
\end{tabular}
}
\end{center}
\caption{Data sets on which the methods are evaluated.}
\label{tab:datasets}
\end{tablehere}
\end{comment}

Testing was performed by cross-validation. Each method was trained on data taken
from chromosomes 1--8 of the genome from which the reads were mapped (including
chromosomes 2a and 2b in the Chimpanzee genome). For evaluation, we drew a set
long, highly expressed exons from the remaining chromosomes. In particular, for
each reference sequence, beginning with the set of exons annotated by Ensembl
release 60 \cite{Hubbard2009}, we removed any exons with known alternate splice
sites, then chose the top 1000 exons by read count, restricting ourselves those
at least 100nt long.

@@

Li recommends that their MART and GLM models be trained using the 100 most abundant
genes. We used 1000 exons from chromosomes 1--8, otherwise chosen in a manner
identical to that which was used to select the test exons.  Both the GLM and
MART model were trained using the default parameters.

Hansen recommends using all of the reads to estimate heptamer frequencies used by
their model. The training procedure works by simple tallying of frequencies and
is thus relatively fast. We choose to use all of the reads from chromosomes 1--8.
We evaluated three variations of this scheme. The first uses the default
parameters, evaluating 7-mer frequencies averaged over the first two positions
of the read (labeled ``Avg-7mer''). We also evaluated estimating 7-mer
frequencies over only the first position, as well as 4-mer frequencies over only
the first position. Though many variations of this scheme may be used, we found these
additional two to be particularly informative. The background frequencies in
each variation were were estimated from positions 18--23 within each read.

All data sets were mapped using Bowtie \cite{Langmead2009} against the latest
genome assemblies available from the UCSC Genome Browser \cite{Karolchik2008}
during preparation.

>>>>

To estimate sequencing bias, we train a Bayesian network classifier in which
each node represents a position in the sequence, relative to the read start, and
edges encode dependency between positions.  Bayesian networks have been applied
to recognize motifs in nucleotide sequences in the past, in particular in
modeling splice sites \citep{Cai2000, Chen2005} and transcription factor binding
sites \citep{Ben-Gal2005, Grau2006, Pudimat2005}.

@@

The model is trained on $n$ sequences, one half labeled as foreground, the other
background, sampled from the reference genome. To obtain the foreground
sequences, we take sequences surrounding the starts positions of $n/2$ aligned
reads. We wish to avoid overfitting to any short locus with a large abundance of reads,
yet we must also capture the bias introduced by PCR amplification. We therefore
choose the $n/2$ unique reads with the highest number of duplicates,
yet count each read only once, ignoring duplicates

To obtain background training sequences, we randomly offset the positions from
which the foreground sequences were sampled.  The offset is drawn from a
zero-mean gaussian, and rounded to the nearest integer, away from zero.
By using such a scheme, we attempt to mitigate the effects of biological
sequence bias, sampling positions that are more likely to be biologically
similar.

This procuder produces a training set of $n$ sequences with accompanying labels
$T = \{ (s_1, x_1), (s_2, x_2), \dots, (s_n, x_n) \}$. The label $x_i$ is
binary, indicating classification as background ($x_i = 0$) or foreground ($x_i
= 1$).

@@

To determine the structure and parameters of the Bayesian network, we use a
hill-climbing approach similar to algorithm described by \citet{Grossman2004}.
The network structure is determined by greedily optimizing the conditional
log-likelihood.
\begin{align*}
\ell &= \sum_{i=1}^{n} \log \Pr( x_i | s_i ) \\
&=
\sum_{i=1}^{n} \log \frac{ \Pr(s_i | x_i) \Pr( x_i ) }{
\sum_{x \in \{0,1\}} \Pr( s_i | x ) \Pr(x) } \\
\end{align*}

@@

As we will be estimating parameters and evaluating the likelihood on the same
set of samples, simply maximizing the likelihood would severely overfit the
training set. We thus penalize model complexity heuristically using the Akaike
information criterion. Where $m$ is the number of parameters needed to specify
the model, we maximize,

$$ \ell' = \ell - m $$

Some benifit would be obtained from a more highly tuned complexity penalty.
However, since the model is trained greedily, additional parameters will be
decreasingly informitive, and increasingly similar between foreground and
background. Adding more parameters will have little effect.  Only when $m$ is
allowed to grow exponentially does the prediction become polluted by small
deviations between thousands of uninformitive parameters.

At each step of the optimization procedure, every possible edge or position
addition, removal, or edge reversal that produces a valid network (i.e., does not
introduce a cycle) is evaluated, and the alteration that increases the score
$\ell'$ the most is kept.  This process is repeated until a local maximum is
found, in which no single alteration to the network will increase the score.

Given the network structure, the parameters are estimated directly from the
observed nucleotide frequencies in the training data.

The run time of the training procedure is further reduced in practice by imposing the
following two restrictions on the structure of the network,
\begin{enumerate}
\item The in-degree (i.e. number of parents) of any node must be less than some
number $p_{\text{max}}$.
\item $|j - i| \le d_{\text{max}}$ for all edges $(i,j)$ and some number $d_{\text{max}}$.
\end{enumerate}
The second of these encodes the assumption that distant nucleotides are
effectively independent. We choose $p_{\text{max}} = 4$ and $d_{\text{max}} =
10$, as reasonable default values.

Figure \ref{fig:models} shows examples of the structure learned when this
procedure is applied to several data sets, using 200,000 reads from each.

@@

Because we cannot observe directly the underlying RNA-abundance, our evaluation
strategy relies on testing two assumptions we make of an ideal, unbiased experiment.
\begin{enumerate}
\item Positional nucleotide frequencies (as in Figure \ref{fig:freqs}), measured
from reads within exons, should not differ greatly from frequencies measured by
sampling uniformly within the same exons.
\item Read counts across a single exon should follow, approxmiately, a Poisson
process.
\end{enumerate}

We applied our procedure (labeled ``BN'') as well as those as well as those of
\citet{Li2010} (``GLM'' and ``MART'') and \citet{Hansen2010} (``7mer''), which
are implemented in the R packages \texttt{mseg} and \texttt{Genominator},
respectively. Thhe procedures were applied to four publically available data
sets \citep{Bullard2010, Mortazavi2008, Trapnell2010, Wetterbom2010}, as well as
an unpublished data sets of our own (Table \ref{tab:datasets}).

\begin{table}
\processtable{Datasets on which the methods are evaluated.\label{tab:datasets}}
{
\begin{tabular}{lllr}\toprule
Experiment & Species & Platform & Read Length \\\midrule
\citet{Wetterbom2010} & Chimpanzee & ABI & 33 \\
Katze (unpublished) & Macaque & ABI & 50 \\
\citet{Bullard2010} & Human & Illumina & 35 \\
\citet{Mortazavi2008} & Mouse & Illumina & 33 \\
\citet{Trapnell2010} & Mouse & Illumina & 75 \\\botrule
\end{tabular}
}{}
\end{table}

Testing was performed by cross-validation. Each method was trained on data taken
from chromosomes 1--8 of the genome from which the reads were mapped (including
chromosomes 2a and 2b of the Chimpanzee genome). For evaluation, we drew a set
of long, highly expressed exons from the remaining chromosomes. In particular, for
each reference sequence, beginning with the set of exons annotated by Ensembl
release 60 \cite{Hubbard2009}, we removed any exons with known alternate splice
sites, then chose the top 1000 exons by read count, restricting ourselves those
at least 100nt long.

@@

\citet{Li2010} recommends that their MART and GLM models be trained using the
100 most abundant genes. We used 1000 exons from chromosomes 1--8, otherwise
chosen in a manner identical to that which was used to select the test exons.
Both the GLM and MART models were trained using the default parameters.

\citet{Hansen2010} recommends using all of the reads to estimate heptamer
frequencies used by their model. The training procedure works by simple tallying
of frequencies. The implementation of this model in the Genominator package we
found to use a great deal of memory, and were unable to train with the volume of
data we wished, so we reimplemented the model and trained it on all of the reads
aligned to chromosomes 1-8.

We evaluated several variations of the heptamer model. The suggested method
involved averaging the frequencies of the first two heptamers of each read. Yet,
we found that in every case, this performed worse than simply counting the
frequencies of the initial heptamer, and thus we report only the latter. The
background frequences are estimated from positions 18--23 in each read.

All data sets were mapped using Bowtie \citep{Langmead2009} against the latest
genome assemblies available from the UCSC Genome Browser \citep{Karolchik2008}
during preparation.

%% leaving off here for now
%% --------------------------------------------------------
